---
output:
  word_document: default
  html_document: default
  pdf_document: default
---
1) Background:
Over the course of the 2019 Fall semester, the students of Nancy Emery's Evolutionary Ecology class (EBIO 4450) conducted and experiment where they subjected a plethora of l. fremontii plants to varying forms for environmental stress. This study was conducted at the CU Greenhouse on 30th Street. Three different experiments were run for this class, all of them looking at how the manipulation of environmental parameters affected plant biomass, the ratio of dispersive to non-dispersive seeds produced by the maternal plant, and the height of the inflorescence, measured from the base of the plant. One experiment, called the "shade" experiment, subjected groups of plants to different levels of light. Another experiment, the "density" experiment, planted 1, 2, or three plants per cone as a measure of competition. The final experiment, the "resource" experiment, subjected plants to varying degrees of available nutrients, another measure of environmental stress. Data was collected near the end of the semester, and each experimental team analyzed their data, with help from Dr. Emery and those that worked in her lab. After the semester ended, as a member of Dr. Emery's lab, was tasked with performing a meta analysis of all the data collected from each experiment, and performing more rigorous testing than we had time for during the class itself. The following is my work!

Experimental Assumptions: Really, the only assumptions made for these experiments, and that is that plant biomass is a measure of stress. This assumption, however, is supported by cited literature. The use of inflorescence height was statistically tested for as a dispersal trait by the class, and the seed type proportion as a measure of dispersal propensity is widely supported by cited literature.

2) Hypotheses

a: Nutrient concentration will have an effect on the plant biomass, the proportion or dispersive to non-dispersive seeds, and inflorescence height.

b: The number of plants in a cone will have an effect on plant biomass, the proportion or dispersive to non-dispersive seeds, and inflorescence height.

c) Access to light will have an effect on the plant biomass, the proportion or dispersive to non-dispersive seeds, and inflorescence height.

3) Predictions

a) Plant biomass will decrease with increasing nutrient concentration; the ratio of dispersive to non-dispersive seeds will increase with nutrient concentration; focal flower height will increase with nutrient concentration

b) With increasing competition, plant mass will decrease; dispersal propensity (seed ratio) will increase with competition; focal flower height will increase with competition

c) Plant biomass will decrease with less access to light; dispersal propensity will be inversely associated with light level; focal flower height will be inversely associated with light level. 


4a) Package loading

```{r}
####  Packages ####
#+ message = FALSE, warning = FALSE
library(tidyverse)
library(nlme)
library(lattice) # contains function 'qqmath' for making qqplot for lmer model
library(lmtest)
# library(glmmTMB)
# library(sjPlot)
# library(sjmisc)
library(knitr)
library("grid")
library("gridExtra")
library(wesanderson)
```

4b) Data processing

```{r}
#'Loading the Data
dat <- read_csv("~/Repos/EmeryLab/DispersalMasterData.csv")

#'Writing conditional to check for NA's in either disk seed count column and filtering accordingly
d2NA <- which(is.na(dat$diskSeedCount2) & !is.na(dat$diskSeedCount1))
d1NA <- which(!is.na(dat$diskSeedCount2) & is.na(dat$diskSeedCount1))
view(dat[d2NA,]) #no cases where d1 is NA and d2 is not, so subset the master data where disk seed count 2 is an NA and disk seed count 1 is not



#'#### Data Cleaning
#' 
#'Getting Mean disk seed count
dat$meanDiskCount <-((dat$diskSeedCount1) + (dat$diskSeedCount2))/2

#replacing observations where disk count 2 was an NA with the value of disk count 1 for meanDiskCount
dat[d2NA,"meanDiskCount"] <- dat$diskSeedCount1[d2NA]
#'Ratio of ray seeds to disk seeds
dat$phyllToTotal <- round(dat$phyllaryCount / (dat$phyllaryCount + dat$meanDiskCount), 3)

#Setting plotting label order for all experiments
labelOrders <- c("Control", "Structure", "Low", "Medium", "High")

```

4c) Subsetting by experiment

```{r}
#### Data Cleaning: Shade Experiment ####
#'

#' #### Data
shadeDat <- dat %>% 
  filter( experiment == "shade")

#' Conditional filtering
shadeDat <- shadeDat %>% 
  filter(!is.na(phyllToTotal)) %>%
  mutate_at(vars(treatmentCat), factor) %>%
  filter(!is.na(treatmentCat))
  

#' Replicates per treatment
shade_reps <- summarize(group_by(shadeDat, treatmentCat), n())
shade_reps

#' Boxplot of the phyllary proportion by treatment
ggplot(shadeDat, aes(treatmentCat, phyllToTotal)) +
  geom_boxplot()

#' Boxplot of the floral height by treatment
ggplot(shadeDat, aes(treatmentCat, focalFlowerHeight_cm)) +
  geom_boxplot()

#' Boxplot of the plant mass by treatment
ggplot(shadeDat, aes(treatmentCat, focalPlantMass_g)) +
  geom_boxplot()
```

```{r}
#### Data Cleaning: Density Experiment ####
#' 
#' #### Data 
densityDat <- dat %>% 
  filter( experiment == "density")

#' Conditional Filtering

densityDat <- densityDat %>% 
  filter( !is.na(phyllToTotal)) %>%
  mutate_at(vars(treatmentCat), factor) %>%
  filter(!is.na(treatmentCat)) %>%
  filter(survivorship == 1)

#' Replicates per treatment
density_reps <- summarize(group_by(densityDat, treatmentCat), n())
density_reps

#' Boxplot of the phyllary proportion by treatment
ggplot(densityDat, aes(treatmentCat, phyllToTotal)) +
  geom_boxplot()

#' Boxplot of the floral height by treatment
ggplot(densityDat, aes(treatmentCat, focalFlowerHeight_cm)) +
  geom_boxplot()

#' Boxplot of the plant mass by treatment
ggplot(densityDat, aes(treatmentCat, focalPlantMass_g)) +
  geom_boxplot()
```

```{r}
#### Data Cleaning: Resource Experiment ####

#' #### Data
#' 
#' Subsetting the data
resourceDat <- dat %>% 
  filter(experiment == "resources")

#' Conditional Filtering
resourceDat <- resourceDat %>% 
  filter(!is.na(phyllToTotal)) %>%
  filter(!is.na(treatmentCat)) %>%
  filter(survivorship == 1) %>%
  mutate_at(vars(treatmentCat), factor)

#' Replicates per treatment
resource_reps <- summarize(group_by(resourceDat, treatmentCat), n())
resource_reps

#' The "high" treatment was removed due to low replication (3 data points)
resourceDat <- resourceDat %>% 
  filter( treatmentCat != "High" ) %>%
  droplevels()


#' Boxplot of the phyllary proportion by treatment
ggplot(resourceDat, aes(treatmentCat, phyllToTotal)) +
  geom_boxplot()

#' Boxplot of the floral height by treatment
ggplot(resourceDat, aes(treatmentCat, focalFlowerHeight_cm)) +
  geom_boxplot()

#' Boxplot of the plant mass by treatment
ggplot(resourceDat, aes(treatmentCat, focalPlantMass_g)) +
  geom_boxplot()
```

5) Shade Experiment: Plant Mass

```{r}
#### Plant Mass: Shade to Control####
#' 
#' Get rid of NA's
shadeDat_mass <- shadeDat %>% filter( !is.na(focalPlantMass_g))

#' __Linear mixed model__
lme_mass_shade <- lme(focalPlantMass_g~treatmentCat, 
                      random = ~1|bin, 
                      data = shadeDat_mass)
summary(lme_mass_shade)
```
*According to this model, which incorporates bin as a potential random experimental design variable, there is no significant difference in plant mass among our different treatment categories*

```{r}
#' _Diagnostic plots: Shade mass_
#' 
#' qqplot to test for normality
qqnorm(lme_mass_shade, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_mass_shade)
```
*Though the residual variances are somewhat different per fitted value, they aren't all that concerning Even so, I fit an unequal variances model and then compared it to the equal variances model.*
```{r}
#' __Linear mixed model with unequal variances__
lme_mass_shade_uv <- lme(focalPlantMass_g~treatmentCat,
                         data = shadeDat_mass,
                         random = ~1|bin,
                         weights = varIdent(form = ~1|treatmentCat))

summary(lme_mass_shade_uv)


#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_mass_shade_uv, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_mass_shade_uv)
```
*The uneuqual variances model to seem to homogenize residual variances, but is it worth it to use this model over the equal variances model at the cost of less degrees of freedom?*

```{r}
anova(lme_mass_shade, lme_mass_shade_uv) # No
```

```{r}
#' Fit the same model without at intercept to get the standard errors of the means for each group (for plotting)
lme_mass_shade_noint <- lme(focalPlantMass_g~ 0 +treatmentCat,
                            data = shadeDat_mass,
                            random = ~1|bin)
```

```{r}
#' _Final answer_
summary_lme_mass_shade <- summary(lme_mass_shade)
summary_lme_mass_shade_noint <- summary(lme_mass_shade_noint)
anova(summary_lme_mass_shade) 
```
*As we can see for this model, there is no significant treatment effect*
```{r}
summary_lme_mass_shade 
```

```{r}
#' Plot the model outcome (work in progress)
shade_mass_output_df <- data.frame(cat_mean_noint = summary_lme_mass_shade_noint$tTable[,"Value"],
                                   cat_se_noint = summary_lme_mass_shade_noint$tTable[,"Std.Error"],
                                   trt_cat = levels(shadeDat$treatmentCat))

shade_mass_output_df

p_shade_mass_output <- ggplot(shade_mass_output_df, aes(x=factor(trt_cat, levels = labelOrders), y=cat_mean_noint)) + 
  geom_errorbar(aes(ymin=cat_mean_noint-cat_se_noint, ymax=cat_mean_noint+cat_se_noint), color = "brown4", width = 0.5) +
  geom_point() +
  geom_jitter(data = shadeDat, aes(x = treatmentCat, y = focalPlantMass_g), color = "brown4", width = 0.1, alpha = 0.4) +
  theme_light() +
  ylab("Focal Plant Mass (g)") +
  xlab("Shade Treatment (µmol of photons/m^2s)") +
  ggtitle("Shade Experiment") +
  theme(plot.title = element_text(hjust = 0.5, vjust = 0.3)) +
  scale_x_discrete(labels = c("Control\n122.0", "Structure\n107.0", "Low\n87.2", "Medium\n71.6", "High\n69.0")) +
  ylim(0,.5) +
  labs(caption = "No Significant Treatment Effect")
p_shade_mass_output
```
*No need for subsequent pairwise comparisons given that the treatment effect was insignificant*

6) Density Experiment: Plant Mass

```{r}
#' Get rid of NA's
densityDat_mass <- densityDat %>% filter( !is.na(focalPlantMass_g))

#' __Linear mixed model__
lme_mass_density <- lme(focalPlantMass_g~treatmentCat, 
                        random = ~1|bin, 
                        data = densityDat_mass)
summary(lme_mass_density)

#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_mass_density, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_mass_density)
```
*Though the residual variances are somewhat different per fitted value, but they aren't all that concerning. Even so, I fit an unequal variances model and then compared it to the equal variances model.*

```{r}
#' __Linear mixed model with unequal variances__
lme_mass_density_uv <- lme(focalPlantMass_g~treatmentCat,
                           data = densityDat_mass,
                           random = ~1|bin,
                           weights = varIdent(form = ~1|treatmentCat))

summary(lme_mass_density_uv)
#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_mass_density_uv, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_mass_density_uv)
```

*The variance structure by fitted value seems to be somewhat more normal using the unequal variances model, though it looks like the gains are marginal, at best. Is it worth it to use the unequal variances model at the cost of degrees of freedom?*

```{r}
anova(lme_mass_density, lme_mass_density_uv) 
```
*Yes, it is!*

```{r}
#' Fit the same model without at intercept to get the standard errors of the means for each group (for plotting)
lme_mass_density_noint <- lme(focalPlantMass_g~ 0 +treatmentCat,
                              data = densityDat_mass,
                              random = ~1|bin,
                              weights = varIdent(form = ~1|treatmentCat))

summary(lme_mass_density_noint)

#' _Final answer_
summary_lme_mass_density_uv <- summary(lme_mass_density_uv)
summary_lme_mass_density_noint <- summary(lme_mass_density_noint)
anova(summary_lme_mass_density_uv)
```
*Given that there was no significant treatment effect, pairwise comparisons are not needed*

```{r}
#Final model output
summary_lme_mass_density_uv
```

```{r}
#' Plot the model outcome
density_mass_output_df <- data.frame(cat_mean_noint = summary_lme_mass_density_noint$tTable[,"Value"],
                                     cat_se_noint = summary_lme_mass_density_noint$tTable[,"Std.Error"],
                                     trt_cat = levels(densityDat$treatmentCat))

density_mass_output_df

p_density_mass_output <- ggplot(density_mass_output_df, aes(x=factor(trt_cat, levels = labelOrders), y=cat_mean_noint)) + 
  geom_point() +
  geom_errorbar(aes(ymin=cat_mean_noint-cat_se_noint, ymax=cat_mean_noint+cat_se_noint), color = "cyan4",width = .5, show.legend = F) +
  geom_jitter(data = densityDat, aes(x = treatmentCat, y = focalPlantMass_g), color = "cyan4", width = 0.1, alpha = 0.4) +
  theme_light() + 
  ylab("Focal Plant Mass (g)") +
  xlab("Density Treatment (plants per cone)") +
  ggtitle("Density Experiment") +
  theme(plot.title = element_text(hjust = 0.5, vjust = 0.3)) +
  scale_x_discrete(labels = c("Control\n0", "Low\n1", "High\n2")) +
  ylim(0,.6) +
  labs(caption = "No Significant Treatment Effect")

p_density_mass_output
```
7) Resource Experiment: Plant Mass

```{r}
#' Get rid of NAs in the data
resourceDat_mass <- resourceDat %>% filter( focalPlantMass_g > 0)

#' __Linear mixed model__
lme_mass_resource <- lme(focalPlantMass_g~treatmentCat, 
                         random = ~1|bin,
                         data = resourceDat_mass)
summary(lme_mass_resource)

#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_mass_resource, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_mass_resource)
```

#' There aren't super clear differences in the variance of the residuals among by fitted value, but I decided to fit the unequal variances model anyway to see what it looked like/if it was worth using.
```{r} 
#' __Linear mixed model with unequal variances__
lme_mass_resource_uv <- lme(focalPlantMass_g~treatmentCat,
                            data = resourceDat_mass,
                            random = ~1|bin,
                            weights = varIdent(form = ~1|treatmentCat))

summary(lme_mass_resource_uv)


#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_mass_resource_uv, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_mass_resource_uv)
```

*Looks like the unequal variances model fixes the issues. But is it worth the extra spending of degrees of freedom?*
```{r}
anova(lme_mass_resource, lme_mass_resource_uv)
```
*Yes*

```{r}
#' _Final answer_
summary_lme_mass_resource_uv <- summary(lme_mass_resource_uv)
anova(summary_lme_mass_resource_uv) 
```
*There was a significant treatment effect; subsequent pairwise comparisons will be necessary*
```{r}
summary_lme_mass_resource_uv
```
*Relative to the control, we can see that both the low and medium treatment categories differed significant. We can see that the low treatment category is associated with a 1.2x mass increase relative to the control (t(25) = 3.16, se = 0.12, p < 0.01). This suggests that the added nutrients within the low treatment category were in fact beneficial to the plant. The medium treatment category is associated with a 38% decrease in mass relative to the control (t(25) = -2.25, se = 0.05, p < 0.05). Plants in this category, when using biomass as a proxy of stress, did no fare as well when compared to the control. Pairwise testing below will illustrate how the plant mass in the two treatments differed when compared to one another* 
```{r}
#Plant Mass: Resource Pairwise Comparisons
resDat_mass <- resourceDat_mass %>% filter( !is.na(focalPlantMass_g))
# Medium To All:
res_mass_medbase <- resDat_mass %>% mutate(treatmentCat = relevel(treatmentCat, "Medium"))

lme_mass_res_medbase <- lme(focalPlantMass_g~treatmentCat,
                            data = res_mass_medbase,
                            random = ~1|bin,
                            weights = varIdent(form = ~1|treatmentCat))

summary(lme_mass_res_medbase)
```
*Here, we can see that the low treatment category is associated with 2.6x mass increase relative to the medium treatment category (t(25) = 4.26, se = 0.11, p < 0.001). The results suggest that plants in the low treatment category faired the best when compared to both the control and the medium treatment*

```{r}
#' Fit the same model without at intercept to get the standard errors of the means for each group (for plotting)
lme_mass_resource_uv_noint <- lme(focalPlantMass_g~ 0 +treatmentCat,
                                  data = resourceDat_mass,
                                  random = ~1|bin,
                                  weights = varIdent(form = ~1|treatmentCat))

summary(lme_mass_resource_uv_noint)

#' _Final answer_
summary_lme_mass_resource_uv <- summary(lme_mass_resource_uv)
summary_lme_mass_resource_uv_noint <- summary(lme_mass_resource_uv_noint)
summary_lme_mass_resource_uv
```

```{r}
#' Plot the model outcome
mass_resource_output_df <- data.frame(cat_mean_noint = summary_lme_mass_resource_uv_noint$tTable[,"Value"],
                                      cat_se_noint = summary_lme_mass_resource_uv_noint$tTable[,"Std.Error"],
                                      trt_cat = levels(resourceDat$treatmentCat))
levels(resourceDat$treatmentCat)
mass_resource_output_df

p_res_mass_output <- ggplot(mass_resource_output_df, aes(x=factor(trt_cat, levels = labelOrders), y=cat_mean_noint)) + 
  geom_point() +
  geom_errorbar(aes(ymin=cat_mean_noint-cat_se_noint, ymax=cat_mean_noint+cat_se_noint),color = "goldenrod3", width = .5, show.legend = F) +
  geom_jitter(data = resourceDat, aes(x = treatmentCat, y = focalPlantMass_g),color = "goldenrod3", width = 0.1, alpha = 0.4, show.legend = F) +
  theme_light() + 
  ylab("Focal Plant mass (g)") +
  xlab("Nutrient Addition Treatment (tbsp nutrients/gal. water)") +
  ggtitle("Eutrophication Experiment") +
  theme(plot.title = element_text(hjust = 0.5, vjust = 0.3)) +
  scale_x_discrete(labels = c("Control\n0", "Low\n0.5x", "High\n1x")) +
  ylim(-0.5, 2) +
  annotate("text", 1, 1, label = "a") +
  annotate("text", 2, 1.75, label = "b") +
  annotate("text", 3, .5, label = "c")  +
  labs(caption = "Significant Overall Treatment Effect; letters indicated within-group differences")
  

p_res_mass_output
```
8) Shade Experiment: Ray Seed Proportion

```{r}
#### Ray Seed Proportion: Shade to Control ####
#' __Linear mixed model__
lme_propn_shade <- lme(phyllToTotal~treatmentCat, 
                       random = ~1|bin, 
                       data = shadeDat)


#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_propn_shade, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_propn_shade)
```
*There are again clear differences in the variance of the residuals among the treatments, so I fit an unequal variances model*

```{r}

#' __Linear mixed model with unequal variances__
lme_propn_shade_uv <- lme(phyllToTotal~treatmentCat,
                          data = shadeDat,
                          random = ~1|bin,
                          weights = varIdent(form = ~1|treatmentCat))


#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_propn_shade_uv, ~ resid(., type = "p"), abline = c(0, 1))
```
*This looks better*
```{r}
#' Residuals vs fitted plot
plot.lme(lme_propn_shade_uv)
```
*This looks better, too. Looks like the unequal variances model fixes the issues. But is it worth the extra spending of degrees of freedom?*
```{r}
anova(lme_propn_shade, lme_propn_shade_uv)
```
*yes!*

```{r}
#Checking for overall treatment effect
anova(lme_propn_shade_uv)
```
*No significant treatment effect; subsequent pairwise comparisons are not necessary*

```{r}

#' _Final answer_
summary_lme_propn_shade_uv <- summary(lme_propn_shade_uv)
summary_lme_propn_shade_uv
```

```{r}
#' Fit the same model without at intercept to get the standard errors of the means for each group (for plotting)
lme_propn_shade_uv_noint <- lme(phyllToTotal~ 0 +treatmentCat,
                                data = shadeDat,
                                random = ~1|bin,
                                weights = varIdent(form = ~1|treatmentCat))

summary(lme_propn_shade_uv_noint)
summary_lme_propn_shade_uv_noint <- summary(lme_propn_shade_uv_noint)


#' Plot the model outcome
shade_propn_output_df <- data.frame(cat_mean_noint = summary_lme_propn_shade_uv_noint$tTable[,"Value"],
                                    cat_se_noint = summary_lme_propn_shade_uv_noint$tTable[,"Std.Error"],
                                    trt_cat = levels(shadeDat$treatmentCat))

shade_propn_output_df

p_shade_propn_output <- ggplot(shade_propn_output_df, aes(x=factor(trt_cat, levels = labelOrders), y=cat_mean_noint)) + 
  geom_point() +
  geom_errorbar(aes(ymin=cat_mean_noint-cat_se_noint, ymax=cat_mean_noint+cat_se_noint), color = "brown4", width = .5, show.legend = F) +
  geom_jitter(data = shadeDat, aes(x = treatmentCat, y = phyllToTotal), color = "brown4", width = 0.1, alpha = 0.4, show.legend = F) +
  theme_light() + 
  ylab("# Phyllaries / Phyllaries + Disk Seeds") +
  xlab("Shade Treatment (µmol of photons/m^2s)") +
  ggtitle("Shade Experiment") +
  theme(plot.title = element_text(hjust = 0.5, vjust = 0.3)) +
  scale_x_discrete(labels = c("Control\n122.0", "Structure\n107.0", "Low\n87.2", "Medium\n71.6", "High\n69.0")) +
  ylim(-.005, .2) +
  labs(caption = "No Significant Treatmne Effect")

p_shade_propn_output

```
8) Density Experiment: Ray seed proportion

```{r}
#### Ray Seed Proportion: Density to Control ####
#' __Linear mixed model__
lme_propn_dens <- lme(phyllToTotal~treatmentCat, 
                      random = ~1|bin, 
                      data = densityDat)
#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_propn_dens, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_propn_dens)
```

*Residual variances look somewhat heteroskedastic, so I fit an unequal variances model*

```{r}
#' 
#' __Linear mixed model with unequal variances__
lme_propn_dens_uv <- lme(phyllToTotal~treatmentCat,
                         data = densityDat,
                         random = ~1|bin,
                         weights = varIdent(form = ~1|treatmentCat))

summary(lme_propn_dens_uv)

#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_propn_dens_uv, ~ residuals(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_propn_dens_uv)
```
*The unequal variances model doesn't seem to change much, but if it is worth using than I will use it*
```{r}
anova(lme_propn_dens, lme_propn_dens_uv)
```
 *The model is not worth using. However, given that neither of these models successfully address the underlying heteroskedasticity, I will use the simpler model, or the equal variances model*
```{r}
anova(lme_propn_dens)
```
*no significant treatment effect; no pairwise comparisons necessary*
```{r}
#Final Answer
summary_lme_propn_dens <- summary(lme_propn_dens)
summary_lme_propn_dens
```

```{r}
#' Plotting

lme_propn_density_uv_noint <- lme(phyllToTotal~ 0 +treatmentCat,
                                data = densityDat,
                                random = ~1|bin,
                                weights = varIdent(form = ~1|treatmentCat))

summary_lme_propn_density_noint <- summary(lme_propn_density_uv_noint)

propn_density_output_df <- data.frame(cat_mean_noint = summary_lme_propn_density_noint$tTable[,"Value"],
                                        cat_se_noint = summary_lme_propn_density_noint$tTable[,"Std.Error"],
                                        trt_cat = levels(densityDat$treatmentCat))

p_dens_propn_output <- ggplot(propn_density_output_df, aes(x=factor(trt_cat, levels = labelOrders), y=cat_mean_noint)) + 
  geom_point() +
  geom_errorbar(aes(ymin = cat_mean_noint - cat_se_noint, ymax = cat_mean_noint + cat_se_noint), color = "cyan4", width = .5, show.legend = F) +
  geom_jitter(data = densityDat, aes(x = treatmentCat, y = phyllToTotal), color = "cyan4", width = 0.1, alpha = 0.4, show.legend = F) +
  theme_light() + 
  ylab("# Phyllaries / Phyllaries + Disk Seeds") +
  xlab("Density Treatment (plants per cone)") +
  ggtitle("Density Experiment") +
  theme(plot.title = element_text(hjust = 0.5, vjust = 0.3)) +
  scale_x_discrete(labels = c("Control\n0", "Low\n1", "High\n2")) 


p_dens_propn_output
```
9) Resource Experiment: Dispersal Propensity

```{r}
#### Ray Seed Proportion: Resource to Control ####
#' Get rid of NAs in the data
resourceDat_propn <- resourceDat %>% filter(!is.na(phyllToTotal))

#' __Linear mixed model__
lme_propn_resource <- lme(phyllToTotal~treatmentCat, 
                          random = ~1|bin,
                          data = resourceDat_propn)

#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_propn_resource, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_propn_resource)
```
*These dont look great; will see if an unequal variances model addresses the issue*

```{r}
#' 
#' __Linear mixed model with unequal variances__
lme_propn_resource_uv <- lme(phyllToTotal~treatmentCat,
                             data = resourceDat_propn,
                             random = ~1|bin,
                             weights = varIdent(form = ~1|treatmentCat))
#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_propn_resource_uv, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_propn_resource_uv)
```

*Looks like the unequal variances model fixes the issues. But is it worth the extra spending of degrees of freedom?*

```{r}
anova(lme_propn_resource, lme_propn_resource_uv)
```
*yes!*

```{r}
#' Fit the same model without at intercept to get the standard errors of the means for each group (for plotting)
lme_propn_resource_uv_noint <- lme(phyllToTotal ~ 0 + treatmentCat,
                                   data = resourceDat_propn,
                                   random = ~1|bin,
                                   weights = varIdent(form = ~1|treatmentCat))

summary(lme_propn_resource_uv_noint)

#' _Final answer_
summary_lme_propn_resource_uv <- summary(lme_propn_resource_uv)
summary_lme_propn_resource_uv_noint <- summary(lme_propn_resource_uv_noint)
summary_lme_propn_resource_uv
```
```{r}
#Checking for overall treatment effect
anova(lme_propn_resource_uv)
```
*There was no significant treatment effect; pairwise comparisons not necessary*
```{r}
#' Plot the model outcome
propn_resource_output_df <- data.frame(cat_mean_noint = summary_lme_propn_resource_uv_noint$tTable[,"Value"],
                                       cat_se_noint = summary_lme_propn_resource_uv_noint$tTable[,"Std.Error"],
                                       trt_cat = levels(resourceDat$treatmentCat))
propn_resource_output_df


p_res_propn_output <- ggplot(propn_resource_output_df, aes(x=factor(trt_cat, levels = labelOrders), y=cat_mean_noint)) + 
  geom_point() +
  geom_errorbar(aes(ymin = cat_mean_noint - cat_se_noint, ymax = cat_mean_noint + cat_se_noint), color = "goldenrod4", width = .5, show.legend = F) +
  geom_jitter(data = resourceDat_propn, aes(x = treatmentCat, y = phyllToTotal), color = "goldenrod4", width = 0.1, alpha = 0.4, show.legend = F) +
  theme_light() +
  ylab("# Phyllaries / Phyllaries + Disk Seeds") +
  xlab("Nutrient Addition Treatment (tbsp nutrients/gal. water)") +
  ggtitle("Eutrophication Experiment") +
  theme(plot.title = element_text(hjust = 0.5, vjust = 0.3)) +
  scale_x_discrete(labels = c("Control\n0", "Low\n0.5x", "Medium\n1x")) +
  ylim(-.01, .5) +
  labs(caption= "No Significant treatment effect")

p_res_propn_output
```
10) Shade Experiment: Inflorescence Height

```{r}
#' Get rid of NA's
shadeDat_height <- shadeDat %>% filter( !is.na(focalFlowerHeight_cm))

#' __Linear mixed model__
lme_height_shade <- lme(focalFlowerHeight_cm~treatmentCat, 
                        random = ~1|bin, 
                        data = shadeDat_height)
#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_height_shade, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_height_shade)
```
*These both look pretty good, but we'll see if an unequal variances model makes it any better*
```{r}
#' __Linear mixed model with unequal variances__
lme_height_shade_uv <- lme(focalFlowerHeight_cm~treatmentCat,
                           data = shadeDat_height,
                           random = ~1|bin,
                           weights = varIdent(form = ~1|treatmentCat))
#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_height_shade_uv, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_height_shade_uv)
```
*Looks like the uv model makes the residuals look better, but is it worth using?*
```{r}
anova(lme_height_shade, lme_height_shade_uv) 
```
*No*

```{r}
#' Fit the same model without at intercept to get the standard errors of the means for each group (for plotting)
lme_height_shade_noint <- lme(focalFlowerHeight_cm~ 0 +treatmentCat,
                              data = shadeDat_height,
                              random = ~1|bin)

summary(lme_height_shade_noint)

#' _Final answer_
summary_lme_height_shade <- summary(lme_height_shade)
summary_lme_height_shade_noint <- summary(lme_height_shade_noint)
summary_lme_height_shade
```

```{r}
#Checking for treatment effect
anova(lme_height_shade)
```
*There was no significant treatment effect; no pairwise comparisons needed*
```{r}
#' Plot the model outcome
shade_height_output_df <- data.frame(cat_mean_noint = summary_lme_height_shade_noint$tTable[,"Value"],
                                     cat_se_noint = summary_lme_height_shade_noint$tTable[,"Std.Error"],
                                     trt_cat = levels(shadeDat$treatmentCat))

shade_height_output_df

p_shade_height_output <- ggplot(shade_height_output_df, aes(x=factor(trt_cat, levels = labelOrders), y=cat_mean_noint)) +
  geom_point() +
  geom_errorbar(aes(ymin=cat_mean_noint-cat_se_noint, ymax=cat_mean_noint+cat_se_noint),color = "brown4", width = .5, show.legend = F) +
  geom_jitter(data = shadeDat, aes(x = treatmentCat, y = focalFlowerHeight_cm),color = "brown4", width = 0.1, alpha = 0.4, show.legend = F) +
  theme_light() +
  ylab("Focal Flower Height (cm)") +
  xlab("Shade Treatment (µmol of photons/m^2s)") +
  ggtitle("Shade Experiment") +
  theme(plot.title = element_text(hjust = 0.5, vjust = 0.3)) +
  scale_x_discrete(labels = c("Control\n122.0", "Structure\n107.0", "Low\n87.2", "Medium\n71.6", "High\n69.0")) +
  ylim(-.5, 20) +
  labs(caption = "No signficant treatment effect")

p_shade_height_output
```

11) Density Experiment: Inflorescence Height

```{r}
#' Get rid of NA's
densityDat_height <- densityDat %>% filter( !is.na(focalFlowerHeight_cm))

#' __Linear mixed model__
lme_height_density <- lme(focalFlowerHeight_cm~treatmentCat, 
                          random = ~1|bin, 
                          data = densityDat_height)
#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_height_density, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_height_density)
```
*These both look pretty good. Will a uv model make it better, and be worth using?*
```{r}

#' __Linear mixed model with unequal variances__
lme_height_density_uv <- lme(focalFlowerHeight_cm~treatmentCat,
                             data = densityDat_height,
                             random = ~1|bin,
                             weights = varIdent(form = ~1|treatmentCat))
#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_height_density_uv, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_height_density_uv)
```
*The gains from using a uv model seem marginal, at best. My guess is model comparison will suggest not using the uv model*
```{r}
anova(lme_height_density, lme_height_density_uv) 
```
*Not worth using the uv model, as suspected*

```{r}
#' Fit the same model without at intercept to get the standard errors of the means for each group (for plotting)
lme_height_density_noint <- lme(focalFlowerHeight_cm~ 0 +treatmentCat,
                                data = densityDat_height,
                                random = ~1|bin)

#' _Final answer_
summary_lme_height_density <- summary(lme_height_density)
summary_lme_height_density_noint <- summary(lme_height_density_noint)
summary_lme_height_density
```

```{r}
#Checking for overall treatment effect
anova(lme_height_density)
```
*No significant treatment effect; pairwise comparisons not neeeded*
```{r}
#' Plot the model outcome
density_height_output_df <- data.frame(cat_mean_noint = summary_lme_height_density_noint$tTable[,"Value"],
                                       cat_se_noint = summary_lme_height_density_noint$tTable[,"Std.Error"],
                                       trt_cat = levels(densityDat$treatmentCat))

density_height_output_df

p_density_height_output <- ggplot(density_height_output_df, aes(x=factor(trt_cat, levels = labelOrders), y=cat_mean_noint)) + 
  geom_point() +
  geom_errorbar(aes(ymin=cat_mean_noint-cat_se_noint, ymax=cat_mean_noint+cat_se_noint),color = "cyan4", width = .5, show.legend = F) +
  geom_jitter(data = densityDat, aes(x = treatmentCat, y = focalFlowerHeight_cm),color = "cyan4", width = 0.1, alpha = 0.4, show.legend = F) +
  theme_light() +
  ylab("Focal Flower Height (cm)") +
  xlab("Density Treatment") +
  xlab("Density Treatment (plants per cone)") +
  ggtitle("Density Experiment") +
  theme(plot.title = element_text(hjust = 0.5, vjust = 0.3)) +
  scale_x_discrete(labels = c("Control\n0", "Low\n1", "High\n2")) + 
  labs(caption = "No significant treatment effect")


p_density_height_output
```
12) Resource Experiment: Inflorescence Height

```{r}
#' Get rid of NAs in the data
resourceDat_height <- resourceDat %>% filter( focalFlowerHeight_cm > 0)

#' __Linear mixed model__
lme_height_resource <- lme(focalFlowerHeight_cm~treatmentCat, 
                           random = ~1|bin,
                           data = resourceDat_height)
#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_height_resource, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_height_resource)
```
*These both look okay, but will a uv model be worth using?*
#' 
```{r}
#' __Linear mixed model with unequal variances__
lme_height_resource_uv <- lme(focalFlowerHeight_cm~treatmentCat,
                              data = resourceDat_height,
                              random = ~1|bin,
                              weights = varIdent(form = ~1|treatmentCat))

#' _Diagnostic plots_
#' 
#' qqplot to test for normality
qqnorm(lme_height_resource_uv, ~ resid(., type = "p"), abline = c(0, 1))
```

```{r}
#' Residuals vs fitted plot
plot.lme(lme_height_resource_uv)
```
*Again, gains are marginal*
```{r}
#Comparing model fit
anova(lme_height_resource, lme_height_resource_uv)
```
*The uv model is not worth using*
```{r}
#' Fit the same model without at intercept to get the standard errors of the means for each group (for plotting)
lme_height_resource_noint <- lme(focalFlowerHeight_cm ~ 0 + treatmentCat,
                                 data = resourceDat_height,
                                 random = ~1|bin)

#' _Final answer_
summary_lme_height_resource <- summary(lme_height_resource)
summary_lme_height_resource_noint <- summary(lme_height_resource_noint)
```
```{r}
#Testing for treatment effect
anova(lme_height_resource)
```
*There was a marginally significant treatment effect. Will conduct pairwise comparisons*
```{r}
summary_lme_height_resource
```
*It would appear that the medium treatment category differed signifcantly from the control, with a ~44% decrease in focal flower height (t(22) = -2.57, se = 1.42, p < 0.05). The low treatment category did not differ significantly from the control. However, given that the overall treatment effect was marginal, I cannot make any strong inferences based on this information*
```{r}
#Pairwise comparisons
resDat_h <- resourceDat_height %>% filter( !is.na(focalFlowerHeight_cm))
# High To All:
res_h_medbase <- resDat_h %>% mutate(treatmentCat = relevel(treatmentCat, "Medium"))

lme_h_res_medbase <- lme(focalFlowerHeight_cm~treatmentCat, 
                         random = ~1|bin, 
                         data = res_h_medbase)

summary(lme_h_res_medbase)
```
*The low treatment category did not differ significantly from the medium treatment category*

```{r}
#' Plot the model outcome
height_resource_output_df <- data.frame(cat_mean_noint = summary_lme_height_resource_noint$tTable[,"Value"],
                                        cat_se_noint = summary_lme_height_resource_noint$tTable[,"Std.Error"],
                                        trt_cat = levels(resourceDat$treatmentCat))

height_resource_output_df

p_res_height_output <- ggplot(height_resource_output_df, aes(x=factor(trt_cat, levels = labelOrders), y=cat_mean_noint)) + 
  geom_point() +
  geom_errorbar(aes(ymin = cat_mean_noint - cat_se_noint, ymax = cat_mean_noint + cat_se_noint), color = "goldenrod4", width = .5, show.legend = F) +
  geom_jitter(data = resourceDat_height, aes(x = treatmentCat, y = focalFlowerHeight_cm), color = "goldenrod4", width = 0.1, alpha = 0.4, show.legend = F) +
  theme_light() +
  ylab("Focal Flower Height (cm)") +
  xlab("Nutrient Addition Treatment (tbsp nutrients/gal. water)") +
  ggtitle("Eutrophication Experiment") +
  theme(plot.title = element_text(hjust = 0.5, vjust = 0.3)) +
  scale_x_discrete(labels = c("Control\n0", "Low\n0.5x", "Medium\n1x")) +
  labs(caption = "Marginally Significant Treatment Effect (p = 0.53)")

p_res_height_output
```

Final Plots
```{r}
#### Final Plots ####

#' Biomass Plot

biomassGrid <- grid.arrange(p_shade_mass_output, p_density_mass_output, p_res_mass_output, nrow = 1, top = "Biomass Response to Treatment Effect by Experiment")
```

```{r}
propnHeightGrid <- grid.arrange(p_shade_propn_output,  p_dens_propn_output, p_res_propn_output, p_shade_height_output, p_density_height_output, p_res_height_output, nrow = 2, ncol = 3, top = "Dispersal Propensity and Inflorescence Height Response to Treatment Effect by Experiment")
```
