---
output:
  pdf_document: default
  html_document: default
---
## Focus of the preceding analyses:
While the original study looked primarily at community composition, I want to take a different approach and investigate whether or not the presence of clonal species acts as a form of stress on non-clonal species. I will use plant biomass as a predictor of perceived stress on behalf of the plant; biomass has been associated with a stress response in plants (support), and thus using this metric as a stress measurement has been supported in the past. Though I cannot take this study to the lengths I might like, I will follow it up with a discussion on how selection might act on the biomass of surviving plants within this experiment, and why this is important. 

Below are the hypotheses I will be testing to address the following questions: 1) Does non-clonal plant biomass differ within each fertilizer plot by community? 2) Does clonal biomass have an effect on non-clonal biomass?

### Hypothesis 1) Community type (presence of clonal vs no presence) has an effect on non-clonal plant biomass

*Predictions: non-clonal plant biomass will be lower in the mixed community than in the non-mixed community across all fertilization treatments*

*Biological Reasoning: clonal plants are effective at growing and usurping available resources and out-competing cohabitants* 

### Hypothesis 2) In the mixed community, clonal plant biomass magnitude will have an effect on non-clonal plant biomass magnitude

*Predictions: There will be an overall decline in non-clonal plant biomass as clonal plant biomass increases; this declination will be more pronounced in small-patch fertilizer treatments based on the recruitment and foraging behavior of clonal plant species in such circumstances, as noted by the paper above*

*Biological Reasoning: clonal plants are effective at growing and usurping available resources and out-competing cohabitants.*

1) Loading Required Packages:

```{r}
library(tidyverse)
library(readxl)
library(nlme)
library(lme4)
library(lattice)
library(grid)
library(gridExtra)
library(car)
```

2) Data processing
```{r}
#Loading the data
dat <- read_csv("~/Repos/School/BioStats/Lab6/Richness_and_Biomass_Lab6.csv")

#Converting characters to factors
dat$Community <- as.factor(dat$Community)
dat$FertilizerTreatment <- as.factor(dat$FertilizerTreatment)
dat$Block <- as.factor(dat$Block)
dat$Year <- as.factor(dat$Year)

#filtering by the year 2008
yearDat <- subset(dat, Year == "2008")

#Filtering by community type "mixed"
mixedDat <- yearDat %>%
  filter(Community == "mixed")

#checking strucutre of the data
str(mixedDat)

#Viewing data
view(mixedDat)
```

```{r}
#distribution of response variable
shapiro.test(mixedDat$NCBMS)#shapiro wilks test indicates normally distributed response variable

#visual inspection of response variable
ggplot(mixedDat, aes(sample = NCBMS)) +
  geom_qq() + 
  geom_qq_line()

hist(mixedDat$NCBMS)
```
*All seems good on this front*

4) Modeling

```{r}
#Analyzing the potential effect of block (experimental design) and the interaction effect of Fertilization treatment and clonal plant biomass
m1 <- lm(NCBMS~ClonBMS*FertilizerTreatment + Block, data = mixedDat)
anova(m1)
summary(m1)
plot(m1)
```
*First, and most importantly, it would appear the the experimental design had no significant effect on the outcome of this study. Second, it would appear that Clonal Plant biomass has a significant effect on non-clonal plant biomass (F(17,1) = 18.30, R.sq = 0.75, p < 0.001) As Well, Fertilizer treatment had a significant effect on non-clonal plant biomass when clonal plant biomass was held constant (F(17,3) = 5.43, p < 0.01)*

```{r}
m2 <- lm(NCBMS ~ ClonBMS*FertilizerTreatment, data = mixedDat)
anova(m2)
plot(m2)
summary(m2)
```
*When removing the effects of block from the model, the results do not significantly change; there is a significant main effect seen for both variables, but no significant interaction term*

```{r}
m3 <- lm(NCBMS ~ ClonBMS + FertilizerTreatment, data = mixedDat)
plot(m3)#looks good
anova(m3)
summary(m3)
```
*With our final, most parsimonious model, we can see that clonal plant biomass has a significant negative effect on non-clonal plant biomass when holding the effect of fertilizer treatment constant (F(27,3)) = 14.61, R.sq = 0.51, p < 0.001). As well, when holding the effect of clonal plant biomass constant, non-clonal plant mass differs significantly by fertilizer treatment (F(27,3) = 4.33, p < 0.05). Relative to the control fertilizer treatment, we see our greatest difference in non-clonal plant biomass (t(27) = 3.23, se = 27.90, p < 0.01), while the large- and uniform-patch treatments did not differe significantly from the control.*

*Final Thoughts and Discussion: These findings are of particular interest, as the original study found that, within the uniform and small-patch fertilizer treatment, non-clonal species numbers were the lowest compared to the control, given that vegetative clonal species could more easily forage into new nutrient territory under such spatially heterogenous conditions. To see that, when the study was all said and done, non-clonal plant biomass grew the most (while controlling for the negative effect of clonal plant biomass) in these conditions indicates that some non-clonal plant species were better suited for these, apparently, highly competitive environments. So while biomass may be used as an justifiable indication of stress, when paired with a significant decline in total non-clonal species numbers, we may actually be seeing the beginnings of selection taking place in these conditions, where surviving non-clonal plants are consuming the nutrients available to them with greater success. Below is a beta regression testing this hypothesis*

```{r}
#Beta regression
mixedDat$proportion <- mixedDat$NCBMS/mixedDat$TotalBMS#creating proportion vector
beta <- betareg(proportion ~ FertilizerTreatment + ClonBMS, data = mixedDat)
plot(beta)#looks good
summary(beta)
```
*As we can see, when accounting for the negative effects of clonal-plant biomass on non-clonal plant biomass, in the small-patch fertilizer treatment, non-clonal plant biomass constituted the largest proportion of total plant biomass relative to the control (z(6)) = 3.266, se = 0.302, p < 0.01). These results, to a degree, validate the predictions above.*
5) Plotting the models
```{r}
#Clonal Plant biomass and non-clonal plant biomass
ggplot(mixedDat, aes(ClonBMS, NCBMS)) +
  geom_point()+
  geom_smooth(method = "lm", data = mixedDat) +
  labs(x = "Clonal Plant Biomass (g/m^2)", y = "Non-Clonal Plant Biomass (g/m^2)", title = "Clonal Plant Biomass Affects Non-Clonal Plant Biomass", subtitle  = "Year 2008", caption = "R.Sq = 0.80") +
  theme_minimal()
```

```{r}
#Fertilization Treatment on non-clonal plant biomass
ggplot(mixedDat, aes(FertilizerTreatment, NCBMS)) +
  geom_point(aes(col = FertilizerTreatment)) +
  geom_boxplot(aes(col = FertilizerTreatment)) +
  labs(x = "Fertilizer Treatment", y = "Non-Clonal Plant Biomass (g/m^2)", title = "Fertilizer Treatment Affects Non-Clonal Plant Biomass", subtitle = "Year 2008") + 
  theme_minimal()
```

```{r}
#Fertilization treatment on NCBMS propotion to total plant biomass
ggplot(mixedDat, aes(FertilizerTreatment, proportion)) +
  geom_boxplot(aes(col = FertilizerTreatment)) +
  labs(y = "Non-Clonal / Total Plant Biomass, (g/m^2)", title = "Biomass Porportions by Fertilizer Treatment", subtitle = "Year 2008") +
  theme_minimal()
```
