---
output:
  pdf_document: default
  html_document: default
---
## Focus of the preceding analyses:
While the original study looked primarily at community composition, I want to take a different approach and investigate whether or not the presence of clonal species acts as a form of stress on non-clonal species. I will use plant biomass as a predictor of perceived stress on behalf of the plant; biomass has been associated with a stress response in plants (support), and thus using this metric as a stress measurement has been supported in the past. Though I cannot take this study to the lengths I might like, I will follow it up with a discussion on how selection might act on the biomass of surviving plants within this experiment, and why this is important. 

Below are the hypotheses I will be testing to address the following questions: 1) Does non-clonal plant biomass differ within each fertilizer plot by community? 2) Does clonal biomass have an effect on non-clonal biomass?

### Hypothesis 1) Community type (presence of clonal vs no presence) has an effect on non-clonal plant biomass

*Predictions: non-clonal plant biomass will be lower in the mixed community than in the non-mixed community across all fertilization treatments*

*Biological Reasoning: clonal plants are effective at growing and usurping available resources and out-competing cohabitants* 

### Hypothesis 2) In the mixed community, clonal plant biomass magnitude will have an effect on non-clonal plant biomass magnitude

*Predictions: There will be an overall decline in non-clonal plant biomass as clonal plant biomass increases; this declination will be more pronounced in small-patch fertilizer treatments based on the recruitment and foraging behavior of clonal plant species in such circumstances, as noted by the paper above*

*Biological Reasoning: clonal plants are effective at growing and usurping available resources and out-competing cohabitants.*

Loading Required Packages:

```{r}
library(tidyverse)
library(readxl)
library(nlme)
library(lme4)
library(lattice)
library(grid)
library(gridExtra)
library(car)
```

# Hypothesis 1: Community type and non-clonal plant biomass

1) Data processing

``` {r}
#The data
dat <- read_csv("~/Repos/School/BioStats/Lab6/Richness_and_Biomass_Lab6.csv")

#Parsing
dat$Community <- as.factor(dat$Community)
dat$FertilizerTreatment <- as.factor(dat$FertilizerTreatment)

#Subsetting by fertilization treatment

#Control
controlDat <- dat %>%
  filter( FertilizerTreatment == "CO")
#Uniform Fert. Treatment
unDat <- dat %>%
  filter( FertilizerTreatment == "UN")
#Small Patch treatment
spDat <- dat %>%
  filter( FertilizerTreatment == "SP")
#Large Patch treatment
lpDat <- dat %>%
  filter( FertilizerTreatment == "LP")

blah <- aov(TotalSppNum ~ Community, data = dat)
summary(blah)
```

2) Preliminary Visualizations
```{r}
#Preliminary visualizations

#Quick prelim vis plotting function
prelimVisPlots <- function(d1,t1, d2, t2, d3, t3, d4, t4, d5, t5, ylab) {
  p1 <- ggplot(d1, aes(Community, NCBMS)) +
    geom_boxplot() +
    labs(y = ylab) +
    ggtitle(t1)
  p2 <- ggplot(d2, aes(Community, NCBMS)) +
    geom_boxplot() +
    labs(y = ylab) +
    ggtitle(t2)
  p3 <- ggplot(d3, aes(Community, NCBMS)) +
    geom_boxplot() +
    labs(y = ylab) +
    ggtitle(t3)
  p4 <- ggplot(d4, aes(Community, NCBMS)) +
    geom_boxplot() +
    labs(y = ylab) +
    ggtitle(t4)
  p5 <- ggplot(d5, aes(Community, NCBMS)) +
    geom_boxplot() +
    labs(y = ylab) +
    ggtitle(t5)
  
  gridPlot <- grid.arrange(p1, p2, p3, p4, p5)
  
  return(gridPlot)
}

#Vis
prelimVisPlots(dat, "Exp. Wide", controlDat, "Control", unDat, "Uniform", spDat, "Small Patch", lpDat, "Large Patch", "nc bms (g m-2)")
```
*Graphically, there seems to be a clear difference in non-clonal biomass in each community experiment wide and within each fertilizer treatment*

3) Modeling

```{r}
#creating function to easily run models
modelFunc <- function (Data) {
  t.test(NCBMS ~ Community, data = Data)
}
#creating a no-intercept model for plotting
modelFunc_noint <- function (Data) {
  t.test(NCBMS ~ 0 + Community, data = Data)
}
#creating a function to create a data frame of important stats for plotting
plotDatFunc <- function(test) {
  dataFrame <- data.frame(estimates = test$estimate,
                      std_error = test$stderr)
  return(dataFrame)
}



```

## Experiment Wide: Community
```{r}
#Experiment wide modeling: testing to see if non-clonal plant biomass in each fertilizer treatmebent differs from the control

#model
expLM <- modelFunc(dat)
#diagnostics plot
plot(expLM)
```
*diagnostics seem fine to me*

```{r}
summary_expLM <- summary(expLM)
summary_expLM
```

*The above model suggests that mean non-clonal biomass in the unmixed community is significantly different than that of the mixed community, where the effect size of presence of clonal species is a 91.3g/m^2 decrease in mean non-clonal plant biomass (p<.001, t = 8.0, DF = 1)*

```{r}
#Plotting

#creating data frame with important statistics (mean, se)
explLM_ni <- lm(NCBMS ~ 0 + Community, data = dat)
summary_expLM_ni <- summary(explLM_ni)

expLM_output_df <- data.frame(cat_mean_noint = summary_expLM_ni$coefficients[, "Estimate"],
                                   cat_se_noint = summary_expLM_ni$coefficients[,"Std. Error"],
                                   trt_cat = levels(dat$Community))
#Creating the plot
expHyp1_plot <- ggplot(expLM_output_df, aes(trt_cat, cat_mean_noint)) +
  geom_point() +
  geom_errorbar(aes(ymin=cat_mean_noint-cat_se_noint, ymax=cat_mean_noint+cat_se_noint), width = 0.2) +
  geom_jitter(data = dat, aes(Community, NCBMS), width = .05, alpha = .4) +
  labs(x = "Community",
       y = "Non-Clonal Plant Biomass (g m-2)") +
  ggtitle(label = "Non-Clonal Plant Biomass: Experiment Wide") +
  theme_minimal()
expHyp1_plot

```

## By Treatment: Control; Community
```{r}
#model
controlLM <- lm(NCBMS ~ Community, data = controlDat)
#diagnostics
plot(controlLM)
```
*Diagnostics look fine to me; assumptions of normalcy met*

```{r}
summary_controlLM <- summary(controlLM)
summary_controlLM
```
*the above model suggests that, within the control, non-clonal plant biomass differes significantly between communities; mean non-clonal plant mass in the unmixed community is approximately 52g/m^2 greater than that of the mixed community (p<.01, DF = 2, T = 2.892)*

```{r}
#Plotting

#creating data frame with important statistics (mean, se)
controlLM_ni <- controlLM <- lm(NCBMS ~ 0 + Community, data = controlDat)
summary_controlLM_ni <- summary(controlLM_ni)

controlLM_output_df <- data.frame(cat_mean_noint = summary_controlLM_ni$coefficients[, "Estimate"],
                                   cat_se_noint = summary_controlLM_ni$coefficients[,"Std. Error"],
                                   trt_cat = levels(controlDat$Community))
#Creating the plot
controlHyp1_plot <- ggplot(controlLM_output_df, aes(trt_cat, cat_mean_noint)) +
  geom_point() +
  geom_errorbar(aes(ymin=cat_mean_noint-cat_se_noint, ymax=cat_mean_noint+cat_se_noint), width = 0.2) +
  geom_jitter(data = controlDat, aes(Community, NCBMS), width = .05, alpha = .4) +
  labs(x = "Community",
       y = "Non-Clonal Plant Biomass (g m-2)") +
  ggtitle(label = "Non-Clonal Plant Biomass: Control Treatment") +
  theme_minimal()
controlHyp1_plot

```

## By Treatment: UN; Community 

```{r}
modelFunc("unLP", "summary_unLP", "summary_unLP_noint", unDat)

```

*NEED TO DO: Convert all lm's to t.tests*


# Hypothesis 2:  In the mixed community, clonal plant biomass magnitude will have an effect on non-clonal plant biomass magnitude

1) Data Processing and Parsing:

```{r}
#Loading the data
dat <- read_csv("~/Repos/School/BioStats/Lab6/Richness_and_Biomass_Lab6.csv")

#Converting characters to factors
dat$Community <- as.factor(dat$Community)
dat$FertilizerTreatment <- as.factor(dat$FertilizerTreatment)
dat$Block <- as.factor(dat$Block)
dat$Year <- as.factor(dat$Year)

#filtering by the mixed community type
mixedDat <- dat %>%
  filter(Community == "mixed")

#checking strucutre of the data
str(mixedDat)

#Viewing data
view(mixedDat)



```

2) Preliminary Data Visualization

```{r}

#Vis. of clonal biomass as predictor of non-clonal biomass
ggplot(mixedDat, aes(ClonBMS, NCBMS)) +
  geom_point(aes(shape = Year)) +
  geom_point(aes(col = FertilizerTreatment)) + 
  geom_smooth(method = "loess") +
  labs(x = "Clonal Biomass", y = "Non-Clonal Biomass")

```
*It seems like there may be a relationship here. Further analysis and the inclusion of other variables will likely add depth to this relationship!*

2) Modeling and Diagnostics

```{r}
#distribution of response variable
shapiro.test(mixedDat$NCBMS)#shapiro wilks test indicates normally distributed response variable

#visual inspection of response variable
ggplot(mixedDat, aes(sample = NCBMS)) +
  geom_qq() + 
  geom_qq_line()

hist(mixedDat$NCBMS)

```

*Both a shapiro wilks test and a visual inspection suggest our response variable is distributed normally*

a) Modeling: Addressing the Hypothesis

```{r}
#Basic linear model with NO inclusion of Fertilizer Treatment
basicLM <- lm(NCBMS ~ ClonBMS, data = mixedDat)
summary(basicLM)
```

```{r}
#model Diagnostics
plot(basicLM)
```
*There seems to be slight heteroskedasticity within residual variances; will try log transformation to see if looks any better*

```{r}
#log transformed model and diagnostics
basicLM_log <- lm(log(NCBMS) ~ ClonBMS, data = mixedDat)
shapiro.test(basicLM_log$residuals)
plot(basicLM_log)
hist(basicLM_log$residuals)

```
*This looks much better; will proceed with further additional complications to this model below*


*Basic Model results: A basic linear model suggests that clonal plant biomass has an effect on the non clonal plant biomass within the mixed community*

b) Modeling: inclusion of Fertilizer Treatment in basic model; ANCOVA
```{r}
#Including Fertilizer Treatment in the basic model
fertLM <- lm(log(NCBMS) ~ ClonBMS + FertilizerTreatment, data = mixedDat)
plot(fertLM,2)#qq looks good
hist(fertLM$residuals)#residuals appear normally distributed
shapiro.test(fertLM$residuals)#almost a perfect normal distribution
summary(fertLM)
```
*According to the output of this model, clonal plant biomass had a significant negative effect on non-clonal plant biomass(F(4,59) = 6.24, R2 = 0.30, se = 0.50, p < 0.001. However, there was no discernible significant difference among non-clonal plant biomass by fertilizer treatments relative to the control when controlling for the effects of clonal plant biomass (F(3,60) = 1.67, se = 45.27, p > 0.05). The takeaway from this is that, in essence, regardless of the spatial heterogeneity of resources, the foraging and resource acquisition potential has a significant negative effect on the biomass non-clonal plant species. Based on the findings of CITE, we can infer that this decrease in biomass is associated with an increase in stress, and thus non-clonal plant species that have a higher potential to increase their biomass in such situations may have a selective advantage.*

c) more variables!


```{r}
#Running the same ANCOVA, but accounting for the effects of time (2006-2008)
yearLM <- lm(log(NCBMS) ~ ClonBMS + FertilizerTreatment + Year + FertilizerTreatment:Year, data = mixedDat)
plot(yearLM,2)#qq looks good
hist(yearLM$residuals)#residuals appear normally distributed, with some abnormalities within the first quantile
shapiro.test(yearLM$residuals)#testing suggests normalcy
summary_YearLM <- summary(yearLM)
summary(yearLM)
test <- as.data.frame(summary_YearLM$coefficients)
test$Names <- names(yearLM$coefficients)
view(test)

ggplot(test, aes(x=Names, y=Estimate, label=Names)) + 
  geom_point(stat='identity', fill="black", size=6)  +
  geom_segment(aes(y = 0, 
                   x = Names, 
                   yend = Estimate, 
                   xend = Names), 
               color = "black") +
  labs(title="Graphical Representation of Estimated Model Effects", 
       subtitle="All effect values relative to Control Fertilizer Treatment in the Year 2006",
       x = "Coefficient Name",
       y = "Effect Size Estimate") + 
  coord_flip() +
  theme_minimal()


view(test)
```
*Things get very interesting here. First, the same negative overall effect of clonal plant biomass on non-clonal plant biomass is seen (F(8,55) = 7.203, adj.R2 = 0.44, se = 0.43, p <0.001). However, the effects of fertilizer treatment become apparent in this model. Relative to the control and not accounting for the effects of year, the large-, uniform-, and small-patch fertilizer treatments do not differ significantly (t = 0.96, se = 0.21, p > 0.05; t = 0.53, se = 0.22, p > 0.05, and t = 1.06, se = 0.22, p > 0.05, respectively). All other variables held constant, we can see that going from the year 2006 to 2008 is associated with a large change in the effect of clonal plant biomass on non-clonal plant biomass (effect size = 0.755, t = 3.01, se = 0.25, p < 0.01). However, when we look further into the model output, we see that the effect of clonal plant biomass on non-clonal plant biomass is dependent on both the level of fertilizer treatment and the level of year. Namely, we see a large opposite effect in how clonal plant mass effects non-clonal plant mass in the small-patch fertilizer treatment, relative to the control, during the year of 2008 (t = 2.48, se = 0.33, p < 0.05); that is, the effect is diminished. This suggests that, with time, the strong observed negative effect tapers off slightly. There was similar change in slope in all other fertilizer treatments in 2008, but these changes were not significant. It should be noted that, broadly speaking, the effects of clonal plant biomass on non-clonal plant biomass diminishes, and most of this diminishing effect is seen as both a function of time and the small-patch fertilizer treatment. Given that much of this change in effect is found in the small-patch fertilizer treatment, this would suggest that non-clonal plants in the small-patch fertilizer treatment may have experienced less stress than those in other nutrient treatments. This result is of particular interest, as it may indicate the importance of genetic diversity as it pertains to competition over time. We predicted that, due to the fact that clonal plant species are more effective at foraging across nutrient boundaries when said boundaries are in closer proximity, non-clonal plant species biomass in such circumstances would decrease at a higher rate relative to those in more spread-out nutrient treatments. To see results that indicate that this may not be the case suggests there may be a threshold at which rapid foraging, resource acquisition, and vegetative growth potential is beneficial, especially when such a plant is foraging into microhabitats that contain different plants, and subsequently more phenotypic diversity.


```{r}
#Checking to see if experimental design factor ("Block") has effect on our predictor variable
blockInteraction_lm <- lm(log(NCBMS) ~ ClonBMS + Block + ClonBMS:Block, data = mixedDat)
summary(blockInteraction_lm)
```
*The results of the above model suggest that, in the case of the 8th block, the effect of Clonal Plant biomass on non-clonal plant biomass depends on the level of that block; our observed effect did not differ significantly among other blocks. In the case of the 8th block, clonal plant biomass had a significantly lesser effect on non-clonal plant biomass relative the same effect within other blocks. In order to accommodate for this, we will removed all data points associated with block 8 from our data set*

```{r}
lm_allVars <- lm(log(NCBMS) ~ ClonBMS + Block + FertilizerTreatment + Year + ClonBMS*Block + ClonBMS*FertilizerTreatment + ClonBMS*Year, data = mixedDat)
summary(lm_allVars)
```
*There's a lot to digest here. First thing that stands out the most is that, when holding all other variables constant, the biomass of non-clonal plants increased significantly from 2006 to 2008. This suggests that, as time progressed, our hypothesized effect became less apparent. This is important given that clonal plant biomass significantly decreased in 2008, and this decrease is dependent on both the change in year and the increase in non-clonal plant biomass with time. Second, there did not appear to be any other significant effect or interaction effect between predictors on non-clonal plant biomass, aside from that of clonal plant biomass. This suggests that, of all the predictor variables and their potential interaction effects, clonal plant biomass is the strongest predictor and, aside from year, accounts for a large amount of the variation seen in non-clonal plant biomass. In summary, we can see a significant decrease in non-clonal plant biomass as clonal plant biomass increases, suggesting that the strong foraging and resource acquisition potential of clonal plants in habitats of spatial/resource heterogeneity is a significant form of stress on non-clonal plant species. Though this is not supported by this analysis in any way, shape, or form, given that we saw a higher average non-clonal plant biomass in 2008 relative to 2006, this suggests that these surviving plants experienced less stress and had a relative higher fitness, suggesting that there may be selection at work in such circumstances. Again, no proof, just a thought.*

```{r}
#subsetting data by year
earlyDat <- subset(mixedDat, Year == "2006")
lateDat <- subset(mixedDat, Year == "2008")

#early basic LM
summary(lm(NCBMS ~ ClonBMS, data = earlyDat))

#late basic LM
summary(lm(NCBMS ~ ClonBMS, data = lateDat))

#mixed model

summary(lm(NCBMS ~ Year + ClonBMS + ClonBMS:Year, data = mixedDat))

```