---
output:
  pdf_document: default
  html_document: default
---
## Focus of the preceding analyses:
While the original study looked primarily at community composition, I want to take a different approach and investigate whether or not the presence of clonal species acts as a form of stress on non-clonal species. I will use plant biomass as a predictor of perceived stress on behalf of the plant; biomass has been associated with a stress response in plants (support), and thus using this metric as a stress measurement has been supported in the past. Though I cannot take this study to the lengths I might like, I will follow it up with a discussion on how selection might act on the biomass of surviving plants within this experiment, and why this is important. 

Below are the hypotheses I will be testing to address the following questions: 1) Does non-clonal plant biomass differ within each fertilizer plot by community? 2) Does clonal biomass have an effect on non-clonal biomass?

### Hypothesis 1) Community type (presence of clonal vs no presence) has an effect on non-clonal plant biomass

*Predictions: non-clonal plant biomass will be lower in the mixed community than in the non-mixed community*

*Biological Reasoning: clonal plants are effective at growing and usurping available resources and out-competing cohabitants.* 

### Hypothesis 2) In the mixed community, clonal plant biomass will have an effect on non-clonal plant biomass

*Predictions: There will be an overall decline in non-clonal plant biomass as clonal plant biomass increases; this declination will be more pronounced in small-patch fertilizer treatments based on the recruitment and foraging behavior of clonal plant species in such circumstances, as noted by the paper above*

*Biological Reasoning: clonal plants are effective at growing and usurping available resources and out-competing cohabitants.*

Loading Required Packages:

```{r}
library(tidyverse)
library(readxl)
library(nlme)
library(lme4)
library(lattice)
library(grid)
library(gridExtra)
library(car)
```

# Hypothesis 1: Community type and non-clonal plant biomass

1) Data processing

``` {r}
#The data
dat <- read_csv("~/Repos/School/BioStats/Lab6/Richness_and_Biomass_Lab6.csv")

#Parsing
dat$Community <- as.factor(dat$Community)
dat$FertilizerTreatment <- as.factor(dat$FertilizerTreatment)

#Subsetting by fertilization treatment

#Control
controlDat <- dat %>%
  filter( FertilizerTreatment == "CO")
#Uniform Fert. Treatment
unDat <- dat %>%
  filter( FertilizerTreatment == "UN")
#Small Patch treatment
spDat <- dat %>%
  filter( FertilizerTreatment == "SP")
#Large Patch treatment
lpDat <- dat %>%
  filter( FertilizerTreatment == "LP")
```

2) Preliminary Visualizations
```{r}
#Preliminary visualizations

#Quick prelim vis plotting function
prelimVisPlots <- function(d1,t1, d2, t2, d3, t3, d4, t4, d5, t5, ylab) {
  p1 <- ggplot(d1, aes(Community, NCBMS)) +
    geom_boxplot() +
    labs(y = ylab) +
    ggtitle(t1)
  p2 <- ggplot(d2, aes(Community, NCBMS)) +
    geom_boxplot() +
    labs(y = ylab) +
    ggtitle(t2)
  p3 <- ggplot(d3, aes(Community, NCBMS)) +
    geom_boxplot() +
    labs(y = ylab) +
    ggtitle(t3)
  p4 <- ggplot(d4, aes(Community, NCBMS)) +
    geom_boxplot() +
    labs(y = ylab) +
    ggtitle(t4)
  p5 <- ggplot(d5, aes(Community, NCBMS)) +
    geom_boxplot() +
    labs(y = ylab) +
    ggtitle(t5)
  
  gridPlot <- grid.arrange(p1, p2, p3, p4, p5)
  
  return(gridPlot)
}

#Vis
prelimVisPlots(dat, "Exp. Wide", controlDat, "Control", unDat, "Uniform", spDat, "Small Patch", lpDat, "Large Patch", "nc bms (g m-2)")
```
*Graphically, there seems to be a clear difference in non-clonal biomass in each community experiment wide and within each fertilizer treatment*

3) Modeling

```{r}
#creating function to easily run models
modelFunc <- function (modelName, summaryModellName, summaryNoIntModelName, Data) {
  modelName <- lm(NCBMS ~ Community, data = Data)
  summaryModellName <- summary(modelName)
  summaryNoIntModelName <- summary(lm(NCBMS ~ 0 + Community, data = Data))
  return(summaryModellName)
  return(summaryNoIntModelName)
}
```

## Experiment Wide: Community
```{r}
#Experiment wide modeling: testing to see if non-clonal plant biomass in each fertilizer treatmebent differs from the control

#model
expLM <- lm(NCBMS ~ Community, data = dat)
#diagnostics plot
plot(expLM)
```
*diagnostics seem fine to me*

```{r}
summary_expLM <- summary(expLM)
summary_expLM
```

*The above model suggests that mean non-clonal biomass in the unmixed community is significantly different than that of the mixed community, where the effect size of presence of clonal species is a 91.3g/m^2 decrease in mean non-clonal plant biomass (p<.001, t = 8.0, DF = 1)*

```{r}
#Plotting

#creating data frame with important statistics (mean, se)
explLM_ni <- lm(NCBMS ~ 0 + Community, data = dat)
summary_expLM_ni <- summary(explLM_ni)

expLM_output_df <- data.frame(cat_mean_noint = summary_expLM_ni$coefficients[, "Estimate"],
                                   cat_se_noint = summary_expLM_ni$coefficients[,"Std. Error"],
                                   trt_cat = levels(dat$Community))
#Creating the plot
expHyp1_plot <- ggplot(expLM_output_df, aes(trt_cat, cat_mean_noint)) +
  geom_point() +
  geom_errorbar(aes(ymin=cat_mean_noint-cat_se_noint, ymax=cat_mean_noint+cat_se_noint), width = 0.2) +
  geom_jitter(data = dat, aes(Community, NCBMS), width = .05, alpha = .4) +
  labs(x = "Community",
       y = "Non-Clonal Plant Biomass (g m-2)") +
  ggtitle(label = "Non-Clonal Plant Biomass: Experiment Wide") +
  theme_minimal()
expHyp1_plot

```

## By Treatment: Control; Community
```{r}
#model
controlLM <- lm(NCBMS ~ Community, data = controlDat)
#diagnostics
plot(controlLM)
```
*Diagnostics look fine to me; assumptions of normalcy met*

```{r}
summary_controlLM <- summary(controlLM)
summary_controlLM
```
*the above model suggests that, within the control, non-clonal plant biomass differes significantly between communities; mean non-clonal plant mass in the unmixed community is approximately 52g/m^2 greater than that of the mixed community (p<.01, DF = 2, T = 2.892)*

```{r}
#Plotting

#creating data frame with important statistics (mean, se)
controlLM_ni <- controlLM <- lm(NCBMS ~ 0 + Community, data = controlDat)
summary_controlLM_ni <- summary(controlLM_ni)

controlLM_output_df <- data.frame(cat_mean_noint = summary_controlLM_ni$coefficients[, "Estimate"],
                                   cat_se_noint = summary_controlLM_ni$coefficients[,"Std. Error"],
                                   trt_cat = levels(controlDat$Community))
#Creating the plot
controlHyp1_plot <- ggplot(controlLM_output_df, aes(trt_cat, cat_mean_noint)) +
  geom_point() +
  geom_errorbar(aes(ymin=cat_mean_noint-cat_se_noint, ymax=cat_mean_noint+cat_se_noint), width = 0.2) +
  geom_jitter(data = controlDat, aes(Community, NCBMS), width = .05, alpha = .4) +
  labs(x = "Community",
       y = "Non-Clonal Plant Biomass (g m-2)") +
  ggtitle(label = "Non-Clonal Plant Biomass: Control Treatment") +
  theme_minimal()
controlHyp1_plot

```

## By Treatment: UN; Community 

```{r}
modelFunc("unLP", "summary_unLP", "summary_unLP_noint", unDat)

```

*NEED TO DO: Convert all lm's to t.tests*